Random Matrix Theory Tools for the Analysis of Functional Magnetic Resonance Imaging Examinations

### 2022-11-13 paper/paper/SPIE/article.tex
[WIP]

\documentclass[12pt]{spieman}  % 12pt font required by SPIE;
%\documentclass[a4paper,12pt]{spieman}  % use this instead for A4 paper
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{graphicx}
\graphicspath{ {./figures/} }
\usepackage{setspace}
\usepackage{tocloft}
\usepackage{doi}
\usepackage{csquotes}
\MakeOuterQuote{"}
\usepackage{xr}
\externaldocument[S-]{supplementary}

@@

\section*{Conclusions}
Eigenfeatures clearly have potential for understanding and reducing fMRI
functional connectivity. However, the utility of these features is strongly
dependent on analytic decisions, suggesting considerable caution when
interpreting past and future studies applying RMT to fMRI.

@@

In small-scale physical systems, RMT universalities have been observed in
quantum chaotic systems, complex nuclei, atoms, molecules and disordered
mesoscopic systems
\cite{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004,brodyRandommatrixPhysicsSpectrum1981,beenakkerRandommatrixTheoryQuantum1997,bohigasHigherOrderCorrelationsSpectra1985,wintgenLevelStatisticsQuantized1988,pandeySkewOrthogonalPolynomialsUniversality2001},
and at larger scales, RMT has been applied to atmospheric physics
\cite{santhanamStatisticsAtmosphericCorrelations2001}, stock cross-correlations
\cite{plerouRandomMatrixApproach2002} , social networks
\cite{jalanUncoveringRandomnessSuccess2014}, random networks
\cite{bandyopadhyayUniversalityComplexNetworks2007}, network-formation in
liquids
\cite{sastrySpectralStatisticsInstantaneous2001,matharooSpectralStatisticsQuenched2009}
and amorphous clusters
\cite{sarkarUniversalityVibrationalSpectra2004,matharooVibrationalSpectraAmorphous2005,matharooUniversalityVibrationalSpectra2008}.
Within biological systems, RMT has also been used to successfully model aspects
of  amino acid functional relationships
\cite{bhadolaTargetingFunctionalMotifs2016}, synchrony in epileptic seizures
\cite{osorioPhasesynchronizationRandommatrixBased2011}, and in protein-protein
interactions both in different species
\cite{agrawalQuantifyingRandomnessProtein2014} and breast cancer
\cite{raiRandomnessPreservedPatterns2015}. RMT has also been used to guide
statistical decisions in principal components
analyses\cite{franklinParallelAnalysisMethod1995,
veraartDenoisingDiffusionMRI2016, ulfarssonDimensionEstimationNoisy2008}, and,
even more recently, has provided insights into the behaviors of deep
neural networks\cite{martinImplicitSelfRegularizationDeep2021,
martinPredictingTrendsQuality2021}.

@@

The functional connectivity—the matrix of correlations of
various collections of voxels of an fMRI scan—is \textit{a priori} a useful
representation of the fMRI data. However, the full correlation matrix between
all voxels is itself often computationally infeasible to work with, and thus is
reduced in various ways prior to being used in analyses.

@@

\begin{table}[h!]
\caption{
    ID = Identifier for paper. TR = Time of Repetition (seconds).
    Time = total duration (minutes) of each scan. Dimensions listed as M x N x P,
    indicate P horizontal / axial slices each with dimensions M x N.
}
\label{table:1}
\small
\centering
\begin{tabular}{ l c c c c c c }
\hline
\textbf{Name}    & \textbf{Dimensions}  & \textbf{Voxel size (mm)} & \textbf{TR} & \textbf{Volumes} & \textbf{n\_scans} \\
\hline
Aging     & 74 × 74 × 32   & 3.0 × 3.0 × 4.0 & 2.0  & 300 & 62  \\
Bilingual    & 100 × 100 × 72 & 1.8 × 1.8 × 1.8 & 0.88 & 823 & 90  \\
Depress   & 112 × 112 × 25 & 2.0 × 2.0 × 5.0 & 2.5  & 100 & 72  \\
Learn     & 64 × 64 × 36   & 3.0 × 3.0 × 3.0 & 2.0  & 195 & 432 \\
Osteo     & 64 × 64 × 36   & 3.4 × 3.4 × 3.0 & 2.5  & 300 & 74  \\
Park      & 80 × 80 × 43   & 3.0 × 3.0 × 3.0 & 2.4  & 149 & 552 \\
Attention & 128 × 128 × 70 & 1.5 × 1.5 × 1.5 & 3.0  & 300 & 90  \\
\hline
\end{tabular}
\end{table}

In order not to inundate readers with dataset details, and because this is an
exploratory investigation of RMT which is not committed to any specific
theory\footnote{In fact, datasets and, later, the classification tasks, were
chosen without investigation into any findings or publications associated with
the data, both in order to reduce bias in our decisions, and because the
original authors analytic / statistical approaches and intentions have very
limited relevance to our whole-brain, voxelwise, multiverse predictive
approach.}, we refer the reader to the original publications and/or data
releases when greater detail is needed, and highlight only the most basic
aspects of each dataset here.  Dataset scan parameters and acquisition details
are summarized in Table \ref{table:1}, and sample and subgroup sizes are
summarized in Table \ref{table:2}.

\begin{table}[h!]
\caption{
    ID = Identifier for paper. Subgroup = name of subgroup. Subjects = Number of subjects.
    Task = fMRI task. ANT = Attention Network Task \cite{fanActivationAttentionalNetworks2005}
}
\label{table:2}
\small
\centering
\begin{tabular}{ l r c c c }
\hline
\textbf{ID}  & \textbf{Subgroup}  & \textbf{Subjects}  & \textbf{Task} & \textbf{Scans per Subject} \\
\hline
Aging   &  older             & 28      & resting-state        & 1 \\
        &  younger           & 34      & resting-state        & 1 \\
        &                    &         &                      &   \\
Bilingual  &  bilingual          & 59      & resting-state    & 1 \\
        &  monolingual        & 33      & resting-state       & 1 \\
        &                    &         &                      &   \\
Depress &  depression         & 51      & resting-state       & 1 \\
        &  control            & 21      & resting-state       & 1 \\
        &                    &         &                      &   \\
Learn   &  task              & 24      & learn image sketches & 16\\
        &  rest              & 24      & resting-state        & 2 \\
        &                    &         &                      &   \\
Osteo   &  duloxetine        & 19      & resting-state        & 1 \\
        &  pain              & 37      & resting-state        & 1 \\
        &  nopain            & 20      & resting-state        & 1 \\
        &                    &         &                      &   \\
Park    &  Parkinson's       & 25      & ANT                  & 12\\
        &  control           & 21      & ANT                  & 12\\
        &                    &         &                      &   \\
Attention&  vigilant           & 11     & resting-state       & 2 \\
        &  nonvigilant         & 11     & resting-state        & 2 \\
        &  trait-attentive     & 11    & resting-state        & 2 \\
        &  trait-non-attentive & 11    & resting-state        & 2 \\
        &  task-attentive     & 11     & resting-state        & 2 \\
        &  task-nonattentive  & 11     & resting-state        & 2 \\
\hline
\end{tabular}
\end{table}

@@

The time and space complexities of computing the full \(N \times N\)
correlation matrix of \(\mathbf{M}\) (and subsequent eigenvalues) are too
large to be tractable. However, since transposition does not change the
eigenvalues, we can use the transpose to efficiently compute these eigenvalues
(see Appendix \ref{sec:transpose}). These raw eigenvalues (feature \code{eigs}
in Table \ref{tab:features}), most directly test the idea that eigenvalues
alone are useful functional connectivity features.

@@

For combined features, slicing variants are also computed, with slicing
performed first on each feature to be combined. E.g. if we have features \(f_1,
\dots,  f_n\) with dimensions \(p_1, \dots, p_n\), then the sliced max 25\%
feature for each component feature, \(\hat{f}_i\) is the last \(p_i / 4\)
elements of \(f_i\), and the combined sliced feature is  \([\hat{f}_1;\;
\dots;\; \hat{f}_n]\)\footnote{Exact implementation can be found in the
\href{https://github.com/DM-Berger/random-matrix-fmri/blob/7c9e4187f582dedee728cd7193b8894d928c2f00/code/rmt/updated\_predict.py\#L182-L213}{source code}}.

@@

The baseline reductions used as statistical summaries were: robust measures of
location (mean, max, min); non-robust measures of location (median,
\(95^{\text{th}}\) percentile, \(5^{\text{th}}\) percentile); non-robust
measures of scale (standard deviation and the range—\code{T-rng} in Table
\ref{tab:features}); and robust measures of scale (interquartile range and
difference between \(95^{\text{th}}\) percentile and \(5^{\text{th}}\)
percentile—\code{T-rrng} in Table \ref{tab:features}). Each statistic reduces
the fMRI data along the voxel dimension, yielding a feature of \(T\) dimensions
(e.g. the ``mean'' feature, \code{T-mean}, is the usual global mean signal).
Exact details can be found in the
\href{https://github.com/DM-Berger/random-matrix-fmri/blob/master/code/rmt/enumerables.py#L51-L69}{paper
source code}.

@@

The AUROC metric was chosen because it handles class imbalances present in the
various datasets, and is naturally normalized and interpretable such that
mAUROC \(<\) 0.5 indicates predictive performance worse than guessing, and
mAUROC \(>\) 0.5 indicates positive predictive
utility\cite{mandrekarReceiverOperatingCharacteristic2010}. However, we also
collected as performance metrics the accuracy, adjusted accuracy (proportion
of samples in the entire dataset largest class minus accuracy), and the F1 score, and a complete
table with these additional metrics is available with the data for this study
(Section \ref{sec:data-and-code}).

Each non-baseline feature is evaluated, for each classification task and
classifier, exactly 1280 times: there are 4 preprocessing levels \(\times\) 2
normalization methods \(\times\) 4 trimming choices \(\times\) 4 unfolding
degrees  \(\times\) 10 slice choices. Each baseline feature is evaluated
exactly 40 times for each comparison task and classifier: 4 preprocessing
levels \(\times\) 2 normalization methods \(\times\) 5 smoothing window sizes.
In total, with all the datasets examined in this study, this yields 2 053 920
mAUROC values to summarize, with 5 main \textit{analytic factors}—
preprocessing, normalization, trimming, unfolding/smoothing degree, and
slicing—to consider.

@@

Summarizing this quantity of data requires some caution. Measures of location or
scale, even when robust, are, for the most part, misleading and uninformative
when distributions are skewed. In most cases we find skewed mAUROC
distributions, and so have chosen primarily to present our findings visually,
with kernel density estimates.

Because aggregating across datasets with different class imbalances can
be extremely misleading, even with a metric like the mAUROC\cite{brabecModelEvaluationNonconstant2020},
we restrict such summaries to Table \ref{tab:numerical} only, and note that
the table is intended as a \textit{supplement only} to the distribution plots: no conclusions
should be drawn about feature performance from the table values alone.

@@

Examining these RMT features more closely, it is clear these features'
performance distributions differ mostly due to the unfolding procedure. That
is, combined features that used the unfolded eigenvalues plus some other RMT
eigenfeature tended to have visually-indistinguishable mAUROC distributions to
those using the unfolded eigenvalues alone (Figures
S\ref{S-fig:fine-trim}-\ref{S-fig:fine-degree}). The mAUROC distributions of
these features differed most strongly in the tails (Figures
S\ref{S-fig:fine-largest}-\ref{S-fig:fine-smallest}).

@@

However, eigenfeature mAUROC values observed in this study were highly
sensitive to the overall analytic procedure, and there was no single analytic
choice (e.g. choice of trimming procedure, unfolding polynomial degree, number
of preprocessing steps, or feature slicing) that ensured, for any combination
of feature and classification task, that all mAUROC values were greater than
0.5. In addition, the mean, median and modal mAUROCs were generally close to
0.5, and adjusted mean and median accuracies also tended to be close to zero.
Thus we find limited evidence that functional-connectivity-based eigenfeatures
have general, "out of the box" predictive utility, with general
utility likely requiring either careful tuning, or different preprocessing
decisions and analytic choices than those examined here.

@@

It is somewhat surprising that the eigenfeatures examined here ever have net
cross-validated predictive utility. The reduction of the functional
connectivity matrix to the sorted \(T - 1\) eigenvalues destroys a large amount
of information (e.g. radically different matrices can have identical sorted
eigenvalues), and the subsequent small-degree polynomial fit used in the
unfolding procedure further reduces variance in the raw data.  All brain voxels
(including grey matter voxels) are also used with equal importance in the
computation of the correlation matrix, and the number of subjects involved in
each dataset is generally quite small.

@@

It was also sometimes the case that adding RMT-specific features (e.g. the spectral
rigidity, or level variance) to either the unfolded eigenvalues, or raw
eigenvalues, would allow for a higher maximum mAUROC (Tables
\ref{tab:numerical}-\ref{tab:top3-fine}, and Figure S\ref{S-fig:fine-largest}). One case where these
features yielded the largest predictions was in the \textit{Aging} dataset.
This is depicted in Figure \ref{fig:observables}, where it is clear that, after
trimming, at almost any \(L\) value, subgroups either cluster in bands, or tend to
generally have lower or higher rigidity of level variance curves.

\begin{figure}
\begin{center}
\begin{tabular}{c}
\includegraphics[width=6.5in]{observables_older_younger_v_older.png}
\end{tabular}
\end{center}
\caption
{ \label{fig:observables} Rigidity and level variance features (unfolding
polynomial degree 3) for \textit{Aging} dataset \footnotesize\texttt{younger v
older} classification task, brain extraction only. Median observed mAUROC =
0.703 (range = 0.424 - 0.880). Horizontal axis L-values range from 1 to 20 in
all subplots, as described in Section \ref{sec:methods}. Note banding / clusters
related to classes.}
\end{figure}

However, the range of mAUROCs observed in all cases in this study suggests
caution. All eigenfeatures and baselines features were sensitive to analytic
choices such as the amount of preprocessing, feature slicing (including trimming) and
feature smoothing / unfolding, and this sensitivity was such that all features
resulted in mAUROCs on either side of 0.5. Most of these analytic choices are benign enough that
it is difficult to rigorously justify one choice over another, meaning RMT-based
analyses may be susceptible to post-hoc justification \cite{kerrHARKingHypothesizingResults1998}.

We believe that future studies using RMT to inform their examinations be more
careful. In the absence of a cross-validated predictive framework that takes
into account the strong impact of researcher analytic choices, any apparent
explanatory utility of RMT should be treated with skepticism. RMT deals with
highly-ideal \textit{iid} matrices, but almost no real-world systems involve
\textit{iid} elements. The basic RMT insight that eigenvalues can be used to
describe matrix-representable systems is certainly vindicated here, but the use
of mathematically non-trivial RMT statistics (like the spectral rigidity or
level number variance) should not be used if simpler statistical descriptions
of the eigenvalues are equally informative.

@@

\section{Disclosures}

\section{Acknowledgments and Funding}

\section{Data, Materials, and Code Availability}
\label{sec:data-and-code}

@@

\subsection{Disclosures}
Conflicts of interest should be declared under a separate header, above
Acknowledgments. If the authors have no competing interests to declare, then a
statement should be included declaring no conflicts of interest. For assistance
generating a disclosure statement, see the form available from  the
International Committee of Medical Journal Editors website:
\linkable{http://www.icmje.org/conflicts-of-interest/}

\subsection{Acknowledgments}
Acknowledgments and funding information should be added after the conclusion,
and before references. Include grant numbers and the full name of the funding
body. The acknowledgments section does not have a section number.

\subsection{Data, Materials, and Code Availability}
Availability of data, materials, and/or code used in the research results
reported in the manuscript may be declared under the heading ``Data, Materials,
and Code Availability,'' following the Acknowledgments section. As relevant,
provide specific access information or restrictions for data, materials, and
computer code (i.e., links to repository access addresses with guidance on
commercial or public access).

@@

\subsubsection{Reference linking and DOIs}
A Digital Object Identifier (DOI) is a unique alphanumeric string assigned to a
digital object, such as a journal article or a book chapter, that provides a
persistent link to its location on the internet. The use of DOIs allows readers
to easily access cited articles. Authors should include the DOI at the end of
each reference in brackets if a DOI is available. See examples at the end of
this manuscript. A free DOI lookup service is available from CrossRef at
\\\linkable{http://www.crossref.org/freeTextQuery/}. The inclusion of DOIs will
facilitate reference linking and is highly recommended.

In the present LaTeX template, the author needs to add the DOI reference by
including it in a ``note'' in the bibliography file, as shown in the file
{\verb+report.bib+}, for example, \\ {\verb+note = "[doi:10.1117/12.154577]"+}.
The DOI may be used by the reader to locate that document with the link:
{\verb+http://dx.doi.org10.1117/12.154577+}.

\subsection{Biographies}
A brief professional biography of approximately 75 words may be provided for
each author, if available. Biographies should be placed at the end of the
paper, after the references. Personal information such as hobbies or
birthplace/birthdate should not be included. Author photographs are not
published.

@@

Denote the correlation matrix of \(\mathbf{X}\) as \(\text{corr}(\mathbf{X})\), and let
\[
\mathbf{Y} = \text{stand}(\mathbf{X}) =
\left(
(\mathbf{X}_1 - \bar{\mathbf{X}}_1) / \sigma_1 | \dots | (\mathbf{X}_p - \bar{\mathbf{X}}_p) / \sigma_p
\right)
\]

>>>>

\documentclass[12pt]{spieman}  % 12pt font required by SPIE;
%\documentclass[a4paper,12pt]{spieman}  % use this instead for A4 paper
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{graphicx}
\graphicspath{ {./figures/} }
\usepackage{setspace}
\usepackage{hyperref}
\usepackage{tocloft}
\usepackage{doi}
\usepackage{csquotes}
\MakeOuterQuote{"}
\usepackage{xr}
\externaldocument[S-]{supplementary}

@@

\section*{Conclusions}
Eigenfeatures clearly have potential for understanding fMRI functional
connectivity in a wide variety of scenarios. However, the utility of these
features is strongly dependent on analytic decisions, suggesting caution when
interpreting past and future studies applying RMT to fMRI.

@@

In small-scale physical systems, RMT universalities have been observed in
quantum chaotic systems, complex nuclei, atoms, molecules and disordered
mesoscopic systems
\cite{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004,brodyRandommatrixPhysicsSpectrum1981,beenakkerRandommatrixTheoryQuantum1997,bohigasHigherOrderCorrelationsSpectra1985,wintgenLevelStatisticsQuantized1988,pandeySkewOrthogonalPolynomialsUniversality2001},
and at larger scales, RMT has been applied to atmospheric physics
\cite{santhanamStatisticsAtmosphericCorrelations2001}, stock cross-correlations
\cite{plerouRandomMatrixApproach2002} , social networks
\cite{jalanUncoveringRandomnessSuccess2014}, random networks
\cite{bandyopadhyayUniversalityComplexNetworks2007}, network-formation in
liquids
\cite{sastrySpectralStatisticsInstantaneous2001,matharooSpectralStatisticsQuenched2009}
and amorphous clusters
\cite{sarkarUniversalityVibrationalSpectra2004,matharooVibrationalSpectraAmorphous2005,matharooUniversalityVibrationalSpectra2008}.
Within biological systems, RMT has also been used to successfully model aspects
of  amino acid functional relationships
\cite{bhadolaTargetingFunctionalMotifs2016}, synchrony in epileptic seizures
\cite{osorioPhasesynchronizationRandommatrixBased2011}, and in protein-protein
interactions both in different species
\cite{agrawalQuantifyingRandomnessProtein2014} and breast cancer
\cite{raiRandomnessPreservedPatterns2015}. RMT has also been used to guide
statistical decisions in principal components
analyses\cite{franklinParallelAnalysisMethod1995,
veraartDenoisingDiffusionMRI2016, ulfarssonDimensionEstimationNoisy2008}, and,
more recently, has provided insights into the behaviors of deep
neural networks\cite{martinImplicitSelfRegularizationDeep2021,
martinPredictingTrendsQuality2021}.

@@

The functional connectivity—often, the matrix of correlations of
various collections of voxels of an fMRI scan—is \textit{a priori} a useful
representation of the fMRI data. However, the full correlation matrix between
all voxels is itself often computationally infeasible to work with, and thus is
reduced in various ways prior to being used in analyses.

@@

\begin{table}[h!]
\caption{
    ID = Identifier for paper. TR = Time of Repetition (seconds).
    Time = total duration (minutes) of each scan. Dimensions listed as M x N x P,
    indicate P horizontal / axial slices each with dimensions M x N.
}
\label{tab:data-dimensions}
\small
\centering
\begin{tabular}{ l c c c c c c }
\hline
\textbf{Name}    & \textbf{Dimensions}  & \textbf{Voxel size (mm)} & \textbf{TR} & \textbf{Volumes} & \textbf{n\_scans} \\
\hline
Aging     & 74 × 74 × 32   & 3.0 × 3.0 × 4.0 & 2.0  & 300 & 62  \\
Bilingual    & 100 × 100 × 72 & 1.8 × 1.8 × 1.8 & 0.88 & 823 & 90  \\
Depress   & 112 × 112 × 25 & 2.0 × 2.0 × 5.0 & 2.5  & 100 & 72  \\
Learn     & 64 × 64 × 36   & 3.0 × 3.0 × 3.0 & 2.0  & 195 & 432 \\
Osteo     & 64 × 64 × 36   & 3.4 × 3.4 × 3.0 & 2.5  & 300 & 74  \\
Park      & 80 × 80 × 43   & 3.0 × 3.0 × 3.0 & 2.4  & 149 & 552 \\
Attention & 128 × 128 × 70 & 1.5 × 1.5 × 1.5 & 3.0  & 300 & 90  \\
\hline
\end{tabular}
\end{table}

In order not to inundate readers with dataset details, and because this is an
exploratory investigation of RMT which is not committed to any specific
theory\footnote{In fact, datasets and, later, the classification tasks, were
chosen without investigation into any findings or publications associated with
the data, both in order to reduce bias in our decisions, and because the
original authors analytic / statistical approaches and intentions have very
limited relevance to our whole-brain, voxelwise, multiverse predictive
approach.}, we refer the reader to the original publications and/or data
releases when greater detail is needed, and highlight only the most basic
aspects of each dataset here.  Dataset scan parameters and acquisition details
are summarized in Table \ref{tab:data-dimensions}, and sample and subgroup sizes are
summarized in Table \ref{tab:data-subjects}.

\begin{table}[h!]
\caption{
    ID = Identifier for paper. Subgroup = name of subgroup. Subjects = Number of subjects.
    Task = fMRI task. ANT = Attention Network Task \cite{fanActivationAttentionalNetworks2005}
}
\label{tab:data-subjects}
\small
\centering
\begin{tabular}{ l r c c c }
\hline
\textbf{ID}  & \textbf{Subgroup}  & \textbf{Subjects}  & \textbf{Task} & \textbf{Scans per Subject} \\
\hline
Aging   &  older             & 28      & resting-state        & 1 \\
        &  younger           & 34      & resting-state        & 1 \\
        &                    &         &                      &   \\
Bilingual  &  bilingual          & 59      & resting-state    & 1 \\
        &  monolingual        & 33      & resting-state       & 1 \\
        &                    &         &                      &   \\
Depress &  depression         & 51      & resting-state       & 1 \\
        &  control            & 21      & resting-state       & 1 \\
        &                    &         &                      &   \\
Learn   &  task              & 24      & learn image sketches & 16\\
        &  rest              & 24      & resting-state        & 2 \\
        &                    &         &                      &   \\
Osteo   &  duloxetine        & 19      & resting-state        & 1 \\
        &  pain              & 37      & resting-state        & 1 \\
        &  nopain            & 20      & resting-state        & 1 \\
        &                    &         &                      &   \\
Park    &  Parkinson's       & 25      & ANT                  & 12\\
        &  control           & 21      & ANT                  & 12\\
        &                    &         &                      &   \\
Attention&  vigilant           & 11     & resting-state       & 2 \\
        &  nonvigilant         & 11     & resting-state        & 2 \\
        &  trait-attentive     & 11    & resting-state        & 2 \\
        &  trait-non-attentive & 11    & resting-state        & 2 \\
        &  task-attentive     & 11     & resting-state        & 2 \\
        &  task-nonattentive  & 11     & resting-state        & 2 \\
\hline
\end{tabular}
\end{table}

@@

The time and space complexities of computing the full \(N \times N\)
\(\mathbf{M}\) of Pearson correlation coefficients (and subsequent eigenvalues) are too
large to be tractable. However, since transposition does not change the
eigenvalues, we can use the transpose to efficiently compute these eigenvalues
(see Appendix \ref{sec:transpose}). These raw eigenvalues (feature \code{eigs}
in Table \ref{tab:features}), most directly test the idea that eigenvalues
alone are useful functional connectivity features.

@@

For combined features, slicing variants are also computed, with slicing
performed first on each feature to be combined. E.g. if we have features \(f_1,
\dots,  f_n\) with dimensions \(p_1, \dots, p_n\), then the sliced max 25\%
feature for each component feature, \(\hat{f}_i\) is the last \(p_i / 4\)
elements of \(f_i\), and the combined sliced feature is  \([\hat{f}_1;\;
\dots;\; \hat{f}_n]\).
% \footnote{Exact implementation can be found in the
%\href{https://github.com/DM-Berger/random-matrix-fmri/blob/7c9e4187f582dedee728cd7193b8894d928c2f00/code/rmt/updated\_predict.py\#L182-L213}{source code}}.

@@

The baseline reductions used as statistical summaries were: robust measures of
location (mean, max, min); non-robust measures of location (median,
\(95^{\text{th}}\) percentile, \(5^{\text{th}}\) percentile); non-robust
measures of scale (standard deviation and the range—\code{T-rng} in Table
\ref{tab:features}); and robust measures of scale (interquartile range and
difference between \(95^{\text{th}}\) percentile and \(5^{\text{th}}\)
percentile—\code{T-rrng} in Table \ref{tab:features}). Each statistic reduces
the fMRI data along the voxel dimension, yielding a feature of \(T\) dimensions
(e.g. the ``mean'' feature, \code{T-mean}, is the usual global mean signal).
% Exact details can be found in the
% \href{https://github.com/DM-Berger/random-matrix-fmri/blob/master/code/rmt/enumerables.py#L51-L69}{paper
% source code}.

@@

The AUROC metric was chosen because it handles class imbalances present in the
various datasets, and is naturally normalized and interpretable such that
mAUROC \(<\) 0.5 indicates predictive performance worse than guessing, and
mAUROC \(>\) 0.5 indicates positive predictive
utility\cite{mandrekarReceiverOperatingCharacteristic2010}. However, we also
collected as performance metrics the accuracy, adjusted accuracy (proportion of
samples in the entire dataset largest class minus accuracy), and the F1 score.
A complete table with these additional metrics is available with the
\hyperref[sec:data-and-code]{data for this study}.

Each non-baseline feature is thus evaluated, for each classification task and
classifier, exactly 1280 times: there are 4 preprocessing levels \(\times\) 2
normalization methods \(\times\) 4 trimming choices \(\times\) 4 unfolding
degrees  \(\times\) 10 slice choices. Each baseline feature is evaluated
exactly 40 times for each comparison task and classifier: 4 preprocessing
levels \(\times\) 2 normalization methods \(\times\) 5 smoothing window sizes.
In total, with all the datasets examined in this study, this yields 2 053 920
mAUROC values to summarize, with 5 main \textit{analytic factors}—
preprocessing, normalization, trimming, unfolding/smoothing degree, and
slicing—to consider.

@@

Summarizing this quantity of data requires some caution. Measures of location
or scale, even when robust, are, for the most part, misleading and
uninformative when distributions are skewed. In most cases we find skewed
mAUROC distributions, and so have chosen primarily to present our findings
visually, with kernel density estimates.  Because aggregating across datasets
with different class imbalances can be extremely misleading, even with a metric
like the mAUROC\cite{brabecModelEvaluationNonconstant2020}, we restrict such
summaries to Table \ref{tab:numerical} only, and note that the table is
intended as a \textit{supplement only} to the distribution plots: no
conclusions should be drawn about feature performance from the table values
alone.

@@

Examining these RMT features more closely, it is clear these features'
performance distributions differ mostly due to the unfolding procedure. That
is, combined features that used the unfolded eigenvalues plus some other RMT
eigenfeature tended to have visually-indistinguishable mAUROC distributions to
those using the unfolded eigenvalues alone (Figures
S\ref{S-fig:fine-trim}-\ref{S-fig:fine-degree}). Instead, the mAUROC
distributions of these features differed mostly in the tails (Figures
S\ref{S-fig:fine-largest}-\ref{S-fig:fine-smallest}).

@@

However, eigenfeature mAUROC values observed in this study were highly
sensitive to the overall analytic procedure, and there was no single analytic
choice (e.g. choice of trimming procedure, unfolding polynomial degree, number
of preprocessing steps, or feature slicing) that ensured, for any combination
of feature and classification task, that all all other analytic choices
resulted in mAUROC values greater than 0.5. In addition, the mean, median and
modal mAUROCs were generally close to 0.5, and adjusted mean and median
accuracies also tended to be close to zero. Thus we find limited evidence that
functional-connectivity-based eigenfeatures have general, "out of the box"
predictive utility, with general utility likely requiring either careful
tuning, or different preprocessing decisions and analytic choices than those
examined here.

@@

It is somewhat surprising that the eigenfeatures examined here ever have net
cross-validated predictive utility. The reduction of the functional
connectivity matrix to the sorted \(T - 1\) eigenvalues uses all brain voxels
(including grey matter voxels), and destroys a large amount of information
(radically different matrices can have identical sorted eigenvalues). The
subsequent small-degree polynomial fit used in the unfolding procedure further
reduces variance in the raw data, and all eigenfeatures, due to the eigenvalue
sorting, are monotonically increasing curves (or approximately monotonic). All
such curves could likely be fit near-perfectly with 3-5 parameters, i.e. the
inherent dimensionality of these features is quite small. Interpreting the eigenvalues
as the magnitudes of the principal components of the standardized data, this
suggests that a rough summary of the magnitudes of the principal components can
often be surprisingly predictive in fMRI.

@@

% It was also sometimes the case that adding RMT-specific features (e.g. the spectral
% rigidity, or level variance) to either the unfolded or raw
% eigenvalues would allow for a higher maximum mAUROC (Tables
% \ref{tab:numerical}-\ref{tab:top3-fine}, and Figure S\ref{S-fig:fine-largest}). One case where these
% features yielded the largest predictions was in the \textit{Aging} dataset.
% This is depicted in Figure \ref{fig:observables}, where it is clear that, after
% trimming, at almost any \(L\) value, subgroups either cluster in bands, or tend to
% generally have lower or higher rigidity of level variance curves.

% \begin{figure}
% \begin{center}
% \begin{tabular}{c}
% \includegraphics[width=6.5in]{observables_older_younger_v_older.png}
% \end{tabular}
% \end{center}
% \caption
% { \label{fig:observables} Rigidity and level variance features (unfolding
% polynomial degree 3) for \textit{Aging} dataset \footnotesize\texttt{younger v
% older} classification task, brain extraction only. Median observed mAUROC =
% 0.703 (range = 0.424 - 0.880). Horizontal axis L-values range from 1 to 20 in
% all subplots, as described in Section \ref{sec:methods}. Note banding / clusters
% related to classes.}
% \end{figure}

\subsection{Limitations}
As is unfortunately typical of fMRI research\cite{turnerSmallSampleSizes2018},
the number of subjects in each dataset was quite small (Table
\ref{tab:data-subjects}). With such small numbers of subjects, randomization
cannot be expected to effectively control group differences, and it is possible
the predictive utility of the eigenfeatures was due to capitalization of such
differences. For example, eigenfeatures were generally predictive in the
\textit{Aging} dataset, and it is quite possible for randomization failures to
introduce age confounding across classes.

Likewise, one of the other more predictable classification tasks was the
\code{duloxetine v nopain} task in the osteoarthritis dataset. An active drug
could introduce any number of physiological confounds relevant to
fMRI\cite{murphyRestingstateFMRIConfounds2013}, but we could not control for
such effects do to the absence of physiological recording in most datasets.

In general, much larger fMRI datasets are needed to adequately control for and test
\textit{what} exactly the connectivity eigenvalues actually predict, and to test
if these predictions generalize to larger, different populations.

@@

Eigenvalue-based features inspired by RMT, and extracted from fMRI functional
connectivity matrices were found to have predictive utility across a wide
variety of datasets and classification tasks. However, the predictive utilities
were modest, and highly dependent on preprocessing steps and other fitting and
feature selection procedures. Future studies employing RMT should carefully
investigate the sensitivity of any findings to such analytic decisions before
drawing any conclusions.

\section{Disclosures}
The authors have no relevant financial interests in the manuscript and no other
potential conflicts of interest to disclose.

% Conflicts of interest should be declared under a separate header, above
% Acknowledgments. If the authors have no competing interests to declare, then a
% statement should be included declaring no conflicts of interest. For assistance
% generating a disclosure statement, see the form available from  the
% International Committee of Medical Journal Editors website:
% \linkable{http://www.icmje.org/conflicts-of-interest/}

\section*{Acknowledgments and Funding}
This work was supported by the Natural Science and Engineering Research
Council of Canada's Canada Research Chair grant (grant number 231266) to JL,
Natural Science and Engineering Research Council of Canada Discovery Grant to
JL, a Canada Foundation for Innovation and Nova Scotia Research and Innovation
Trust infrastructure grant (R0176004) to JL, a St. Francis Xavier University
research startup grant to JL (grant number R0168020), and a St. Francis Xavier
University UCR grant to JL.

\section*{Data, Materials, and Code Availability}
\label{sec:data-and-code}

@@

% \subsection{Disclosures}
% Conflicts of interest should be declared under a separate header, above
% Acknowledgments. If the authors have no competing interests to declare, then a
% statement should be included declaring no conflicts of interest. For assistance
% generating a disclosure statement, see the form available from  the
% International Committee of Medical Journal Editors website:
% \linkable{http://www.icmje.org/conflicts-of-interest/}

% \subsection{Acknowledgments}
% Acknowledgments and funding information should be added after the conclusion,
% and before references. Include grant numbers and the full name of the funding
% body. The acknowledgments section does not have a section number.

% \subsection{Data, Materials, and Code Availability}
% Availability of data, materials, and/or code used in the research results
% reported in the manuscript may be declared under the heading ``Data, Materials,
% and Code Availability,'' following the Acknowledgments section. As relevant,
% provide specific access information or restrictions for data, materials, and
% computer code (i.e., links to repository access addresses with guidance on
% commercial or public access).

@@

% \subsubsection{Reference linking and DOIs}
% A Digital Object Identifier (DOI) is a unique alphanumeric string assigned to a
% digital object, such as a journal article or a book chapter, that provides a
% persistent link to its location on the internet. The use of DOIs allows readers
% to easily access cited articles. Authors should include the DOI at the end of
% each reference in brackets if a DOI is available. See examples at the end of
% this manuscript. A free DOI lookup service is available from CrossRef at
% \\\linkable{http://www.crossref.org/freeTextQuery/}. The inclusion of DOIs will
% facilitate reference linking and is highly recommended.

% In the present LaTeX template, the author needs to add the DOI reference by
% including it in a ``note'' in the bibliography file, as shown in the file
% {\verb+report.bib+}, for example, \\ {\verb+note = "[doi:10.1117/12.154577]"+}.
% The DOI may be used by the reader to locate that document with the link:
% {\verb+http://dx.doi.org10.1117/12.154577+}.

% \subsection{Biographies}
% A brief professional biography of approximately 75 words may be provided for
% each author, if available. Biographies should be placed at the end of the
% paper, after the references. Personal information such as hobbies or
% birthplace/birthdate should not be included. Author photographs are not
% published.

@@

Denote the correlation matrix of \(\mathbf{X}\) as \(\text{corr}(\mathbf{X})\), and let
\[
\mathbf{Y} = \text{standardize}(\mathbf{X}) =
\left(
(\mathbf{X}_1 - \bar{\mathbf{X}}_1) / \sigma_1 | \dots | (\mathbf{X}_p - \bar{\mathbf{X}}_p) / \sigma_p
\right)
\]
